---
title: "Project"
output: html_document
---

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## Analysis

### Getting the data
Need for consideration of diverse 'NA' values in raw files. Usage of data.table package faciliates reading a column class assigment (note that data.table syntax is a bit weird)

```{r}
require(data.table)
require(caret)
### data reading
train.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
train <- fread(train.url)

validation.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
validation <- fread(validation.url)

### explore train set
# str(train)
# --> several NA values

### redo data reading
train.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
train <- fread(train.url, na.strings=c("NA","", "#DIV/0"))
# str(train)

validation.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
validation <- fread(validation.url, na.strings=c("NA",""))
# str(validation)

### explore dimensions
dim(train)
dim(validation)
```

I get rid of non-useful columns, e.g. 95% of NA values
```{r}
### 1. NAs
# get NA values by column
NAbycol <- train[,lapply( .SD, function(x) sum(is.na(x))/nrow(train) )]

# 0.95 seems to be a good threshold
myNAcols <- as.vector(NAbycol>.95)
mycols <- names(NAbycol)[myNAcols]

# and get rid of the columns
train[,(mycols):=NULL]
validation[,(mycols):=NULL]

### 2. nearly zero variance
nz <- nearZeroVar(train)
nz <- names(train)[nz]
train[,(nz):=NULL]
validation[,(nz):=NULL]

```

Therefore, only the following variables will be considered in the analysis:
```{r}
train[, classe:= factor(classe)]
# train
str(train)
```


### Choosing a model

I divide the 'train' data into 'training'(60%) and 'testing'(40%) sets
```{r}
set.seed(12345)
inTrain <- createDataPartition(y=train$classe, p=0.6, list=FALSE)
training <- train[inTrain,]
testing <- train[-inTrain,]
rm(train)
```

I will try first 'radom forest' model, which is known to give highly accurate predictions. The model is 1st fit on the 'training' data (60% of 'train'set ). A 3-fold cross validation is used to select the tuning parameters


#### Random forest

```{r}
#  a 3-fold cross validation will be used to select optimal tuning paarmeters
trCont <- trainControl(method="cv", number=3, verboseIter=F)

# fitting the model on training set
fit <- train(classe ~ ., data=training, method="rf", trControl=trCont)

fit$finalModel

```

'classe'is predicted now in 'testing' set (40% of "train" set). The confusion matrix is shown to compared predicted vs reference values

```{r}
pred <- predict(fit, newdata=testing)
confusionMatrix(training$classe, pred)
```

As expected, we get a high accuracy (99.7%), being then the accuracy for the out-of-sample error of 0.3%
Considering this good result, I am not considering any other model

The model is retrained using the completed 'train'set
```{r}
#  a 3-fold cross validation will be used to select optimal tuning paarmeters
trCont <- trainControl(method="cv", number=3, verboseIter=F)

# fitting the model on training set
fit <- train(classe ~ ., data=train, method="rf", trControl=trCont)
```
